\normalfont\spacedallcaps{The economic consequences of different attitudes of a policy maker: a combined epidemiological-econometric study}

### 2021-03-12 Manuscript/manuscript.tex
PUB

\title{\normalfont\spacedallcaps{The economic consequences of the different attitudes of a policy maker: a combined epidemiological-econometric study}} % The article title

@@

@@

The recent pandemic due to the spread of SARS-CoV-2 virus,  opened a highly debated issue about what it would be the best approach of the policy maker to face the epidemics.  Differently from the pandemics of the past and in particular of the XX century (e.g. Spanish Flu,  Asiatic Flu and Hong-Kong Flu),  for this pandemic a very large amount of data are easily accessible.  As consequence not only the modelling of the virus diffusion,  but also the effect on socio-economic texture for the different countries can be investigated with a finer resolution.  Among the different scientific challenges that can come up in this context,  an interesting one involves the socio-economic effects of the attitude of the policy-maker (PM) to block the circulation of people (lock-down) in order to reduce the contagion rate (more formally the reproduction number, as described in Supporting Information).  Indeed the policy maker can adopt, at a first approximation,  a linear combination between these two extreme approaches: forcing all people to stay at home or to \textit{laissez-faire}.  In the first extreme,  the spread of virus is,  of course, stopped but,  on the other side,  the toll for such approach is that not only the economic activity (and so the income of people/firms) but also that the furniture of the primary goods are stopped. On the other side,  if no lock-down measures are taken by the PM the toll to be paid will be not only the high number of deaths but also the economic damage produced by the very high number of deaths \cite{Correia2020,karlsson2014impact}.  In practice the PM can adopt intermediate approaches that shut down activities that contribute much more to diffusion with respect to other (for this purpose a very fine analysis was provided by Li et al.  in  \cite{li2021temporal} and by Brauner et al. in \cite{brauner2021inferring}).  As consequence the lock-down efficacy,  within certain limits,  can be tuned.  In the literature different scholars \cite{kabir2020evolutionary,rowthorn2020cost} challenged the issue of finding the optimal lock-down policy for minimizing the economic impact as well as the deaths.  In particular for the model in Ref. \cite{kabir2020evolutionary} it is assumed that the policy-maker know perfectly the consequences of her choices and that she can act without delay to impose the optimal choice; finally it is assumed that the PM can impose a continuous factor for the lock-down, while, for different countries such factor seems to be much more discrete.  It can be argued that most of these drawbacks of this last formidable research,  are entangled with the fact that a deterministic approach was considered for the activation of the lock-downs by the policy maker. On this basis we would propose here an alternative way to model the decision of the policy maker that is based on a stochastic model instead of on a deterministic one.  Furthermore, differently with respect to the previous researches, that focused basically on macroeconomic impact,  here we modelled the impact of the different lock-down at microeconomic level: in particular,  by means of difference in difference,  we evaluated the effect of the different levels of lock-down on the sales for different product types in Italy.  Thus the final output of the model will be not only the cumulative deaths,  but also the economic damage for each selling sector. Moreover here we also considered that there is not only an economic cost for each death,  as done by \cite{kabir2020evolutionary} but there is also average cost for each infected person (referring to Italian data),  because a consistent part of them may be recovered or even should take the intensive therapy.  As we will show by varying with continuity the probabilistic parameter by which the PM impose the lock-down,  a discontinuity in the SIRD model and in the days of the lock-down was obtained.  Such behaviour belongs to a class of phenomena that, in physical sciences, is called a phase transition.  For each phase we will discuss the result of the simulation and then we will compare them in order to get a general insight.

@@

The model of the present study is composed by an epidemiological part,  that shapes the diffusion of the virus, and then its output is used by the economic model to quantify the damage. Thus we will discuss the epidemiological part and then the economic one

@@

Among the very large number of compartmental model that are available in the literature \cite{vynnycky2010introduction},  we considered as the simulator of the epidemic diffusion the simplest one: the Susceptible-Infectious-Recovered-Dead (SIRD).  Our choice is motivated by the fact that this relative simple model provide the gross features of an epidemic \cite{fernandez2020estimating,al2020forecasting} with a relative small number of parameters. Thus,  because the final goal of the present article is an insight on the effects of the PM care of pandemic,  and not a fine grained description of the epidemics due to COVID-19,  such model is suitable for our proposes\footnote{One in principle can consider a SIRD model,in which the parameters that are time-dependent, as done by Ferrari et al. in Ref. \cite{ferrari2020modelling} for the description of Italian situation.  On the other side it is possible to increase the complexity of the model with other compartments as done in the following paper \cite{giordano2020modelling} by Giordano et al.  Note that in this last case the resolution of 9 differential equation is required (accompanied by the estimation of a large number of parameters)}). The SIRD model,first proposed by Kermack and McKendrick in 1927 \cite{kermack1927contribution},  is given by the following set of differential equations\cite{vynnycky2010introduction}:  

@@

know as the basic reproduction factor. This number,  that is the average number of people that are infected by a single individual,  describes if the epidemic is in a negative feedback (R$_{0}$<1),  stationary (R$_{0}$=1),  or in a positive feedback (R$_{0}$>1).  As consequence if the epidemic is in a negative feedback it will be going to dissipate,  if it is in positive feedback it will grow.  Note that in this simple model since the parameters are not time dependent this factor is constant.  If,  as performed by Ferrari et al. \cite{ferrari2020modelling},  time dependent parameters are considered a time depended parameter called $R_{t}$.  For the present work,  we limited to constant parameters: in particular we considered the parameter estimation for Lombardy provided by Neves and Guerrero in Ref \cite{neves2020predicting}: $\beta$ was set equal to 0.55 while $\gamma$ equal to $\frac{1}{7}$. The overall population $N$ was set to 60M,  in order to simulate Italian population.  Within the daily temporal evolution of this model, that was obtained by numerically solving the differential equation above via the \textit{DeSolve} package,  we considered a trigger activated by the PM every 7 days: if the number of infected people normalized by the overall population is more than $1\times 10^{-7}$,  there is a probability that the PM impose laws that multiply the $\beta$ factor by $0.7$,  if the normalized infected people are more than 10 the previous threshold  she will impose,  with a certain probability, restrictions that multiply the $\beta$ starting $\beta$ factor by $0.25$,  finally if the  threshold is exceeded more than 50 times the PM will impose with a certain probability restrictive measures that multiply the $\beta$ by $0.025$.  These attenuation parameters were adjusted taking in consideration the results of Marziano et al. in Ref. \cite{marziano2021retrospective}. As we said the PM act with a certain probability, more formally stochastically: each week a random number (from zero to one) is extracted,  if this is higher than a certain threshold, the relative restrictive decision is taken, otherwise not. The threshold value capture the PM attitude to impose the lock-down: lower values model a careful PM,  hight value a lazy one. In this way the model is able to simulate different scenarios for the different PM attitude: as we will see this can produce two very different results.  At the end of the simulation,  beside the values given by the standard SIRD model (recovered and deaths),  the model also provide the number of days in which each restriction was active: these values are then used for the economic model in order to evaluate the economic effect due to the restrictions and PM strategy.  It is worth nothing that here as lock-down we considered only the national one that was applied in the first wave of the epidemics: from March 2020 to June 2020. This choice is motivated by the fact that modelling a unique system is more simpler with respect to modelling an ensemble of a communicating clusters that represent region or provinces: therefore,  if one is interested to model the second wave of pandemics such approach should be undertaken.  Moreover in this latter case,  as a further degree of complexity,  the economic data described in the next subsection must be at regional or province level,  and as far as the authors know, such data are not available.  For these reasons here the modelling will be always referred to national data and national lock-downs,  therefore the period after June when starting from October regional lock-down were imposed,  will be not considered.

@@

The economic impact for each epidemiological scenario is shaped as follow: a first set of parameters,  as the economic value of a death and of being infected by COVID-19 ,was taken directly from the reports/documentation of official sources; other parameters as the effect on selling for different areas and on the unemployment benefit (Cassa Integrazione Guadagni) were evaluated with empirical approaches from raw data.  Concerning the first set,  the number of deaths is multiplied by the maximum compensation value provided by the Court of Milan \cite{tribunaleMilano} for manslaughter (300k EUR). This choice is based on the idea that,  if the PM act improperly,  can be incriminated for manslaughter (with the consent of parliament that has to validate the incrimination) and than,  if judged guilty,  charged by this amount for each death\footnote{In principle the judge also keep into account the age of the deaths: this in principle require an epidemiological model in which also the age of people is keep into account. In this case however a system of partial differential equation should be solved making the calculation and the computational cost incredibly high.}.  Beside this impact, there is also the cost associated to the medical care of each ill people,  for this we considered the average value calculated by National Anti-Corruption Authority (ANAC) \cite{Anac}: 28.180 kEUR \footnote{It is worth noting that,  in principle,  there is also another import health-care impact due to the fact that the ill people for COVID-19 saturate the health system thus making it unreachable for other diseases.  This spillover translates in to more death and more ill people with respect to the baseline situation where there isn't an pandemic: however,  by now,  this effect is difficult to quantify and so we did not included in to the present model}.  Among the different sectors affected by the pandemic and the consequent lock-down,  we focused on the selling for the following ATECO-2007 \cite{Ateco2007} categories\footnote{The category Equipment for information technology/telephony and telecommunications was omitted since there is no point to argue that the lock-down affected, in reverse there it is more likely to argue that is increased due to the fact that people was forced to work remotely \cite{guo2020digitalization}}  \footnote{In the rest part of the paper these categories will be referred as the part of the name labelled in blue. 
}: \textcolor{blue}{Food}, \textcolor{blue}{Clothing} and furs, \textcolor{blue}{Footwear}/leather and travel articles, \textcolor{blue}{Furniture}/textile articles/furnishings for the home, 
\textcolor{blue}{Appliances}/radios/televisions and tape recorders, \textcolor{blue}{Photo-optics}/films/compact discs/ audio-video cassettes and musical instruments, Durable and non-durable \textcolor{blue}{Household kinds}, \textcolor{blue}{Household tools} and hardware,  \textcolor{blue}{Games}/toys/sports and camping articles. The choice to use the selling as a parameter for the evaluation of the lock-down lies on the fact that with them it is possible to capture not only the contraction for each sector but also the loss for the public treasury due to to the reduced incomes from the VAT\footnote{For this purpose another sector that in principle can be also considered is the contraction of fuel selling, due to the reduced mobility, where in addition to the VAT there is also a fixed taxation (accisa).  Such calculation may be considered as a future outlook of this work}.  Beside the selling we also considered the unemployment benefit for the following ATECO sectors: \textcolor{blue}{Manufacturing} activities, \textcolor{blue}{Construction} , \textcolor{blue}{Wholesale} and retail trade/ repair of motor vehicles-motorcycles and personal and household goods.  In this case the choice to use also this parameter is based on the fact that this is the first aid provided by the government for the firms that were damaged by the contraction in demand as well as by the fact that for different sector of them the production was also banned by law during the lock-down \footnote{In principle one can be also interested to disentangle these two effects: in this case an interesting option is to study the second-wave of the epidemics in Italy that happened in the second half of the 2020. Contrary with respect to the first wave, the lock-down did not banned by law the production.  As told previously this option was not considered by the authors since the most of the economic data were available, with the temporal resolution considered in this work, only at national level.}.  For the empirical evaluation of the impact on selling and unemployment benefit we performed a Difference-In-Difference (DID) where multiple times where considered as used in Refs. \cite{draca2011panic} and \cite{wooldridge2012introductory}. The following regression was performed: 

@@

where $Y$ is the selcted outcome, $\alpha$ the intercept, $T_{i}$ a dummy variable for the lock-down timing i,  C a dummy for the treated group and $\epsilon$ a error term.  As control group for the sales we considered the food ones since,  in principle,  people can be considered to use almost the same amount of food regardless with respect to the lock-down \footnote{It is true that a slight increase of food consumption was done by people that were forced to stay at home as the plots in Fig.  \ref{Selling_plots} \ref{Selling_plots_II} captures: however as proven numerically in the Supporting Info this change does not significantly affect the DID estimations } (note that in this selling also the one for the restaurants are included: thus,  if it is true that the restaurants were closed,  it is also true that people will purchase basically the same amount of food from the supermarket that they will consume into restaurant if these were opened).  On the other side,  for subsidies \textit{Cassa integrazione},  we considered the Cassa Integrazione Straordinaria - Solidarità as the control group (note that this subsidy is different with respect Cassa Integrazione Solidarità that was dedicated to sectors not covered by Cassa Integrazione Ordinaria).  This subsidy can be used by firms in order to reduce the working hours of employee by keeping part of the original salary.   However,  during the first months of pandemic firms preferred (as it was indicated by Italian Government),  for reducing the daily hours worked,  to force employee to use holiday days in order to keep their salary constant.  Thus  this particular subsidy can be considered as not treated by the first wave of the lock-down.  On the contrary the Cassa Integrazione Ordinaria is largely affected by the lock-down since firms use this subsidy in order to reduce the monthly costs of employee (in particular when the holiday day of a worker were zero or negative).  The DID considered here has three different times for the lock-down:  for the sales we considered, as  done for the epidemiological model,  the months of March 2020 and May 2020 as medium lock-down, the month of April 2020 as High lock-down and June 2020 as weak lock-down.  On the other side  concerning the subsidies we considered as medium lock-down only May 2020: such choice was motivated by the fact that the effect of firms to use the subsidies for workers was slightly delayed with respect to effect lock-down effects on sales.  Performing the DID we estimated the coefficients for the effect of each intensity of lock-down on sales and subsidies,  then we we rescaled them for a week (multiplying them by a factor of $\frac{7}{30})$. Finally we multiplied the number obtained from the previous calculation by the number of lock-down weeks, with the respective intensity,  in each scenario simulated via the epidemiological model described before.  As consequence we obtained,  for each scenario, an economics simulation based on parameters obtained from an empirical evaluation.

@@

The monthly sales data were take retrieved from the website of National Institute of Statistics \cite{istat}.  In particular we considered the period starting from 2018 up to June 2020. The choice to not consider the month after June and in particular the last part of the year lies on the fact that in the latter the lock-down were imposed at regional level and not to national level \footnote{With the only exception of Christmas holiday}  \footnote{In principle one may ask why the present analysis was not performed on regional cluster making it more flexible: unfortunately the economic data used here,  
as far as the author knows, were not available,  at all for regional cluster. Moreover if regional cluster were considered it was necessary to model an ensemble of SIRD model that communicate each other with a defined rate (that change also with respect to the lock-down restriction).  This make the model much more complicated.  However if all the data that are necessary for performing the analysis will become availbale the author may consider, as an outlook, to extend the present analysis to regional clusters} As discussed in Supporting Information,  these data were not de-seasoned: thus we performed a de-seasoning via the \textit{Forecast} R package \cite{forecast} that uses a Hilbert-Huang transform  \cite{huang1998empirical} for the decomposition of a time series data.  Concerning the subsidies we retried the data from the \textit{Osservatorio Cassa integrazione guadagni e fondi di solidarietà} on the \textit{Istituto nazionale della previdenza sociale} webpage \cite{inps}.  It is worth nothing that here we considered only the authorized subsidy and not the asked one.  Furthermore these data,  differently with respect to the sales one were not affected by seasonality noise,  and thus no de-seasoning was necessary.

@@

As done for the section Model and methods we will divide the discussion of the result in the following way: first the result of the economic model will be presented,  then on the basis of them we will discuss the  scenarios obtained with them via the epidemiological model.  Finally we will discuss the overall results.

@@

For each sales category we run the regression reported in the Eq.  \ref{DID-eq} on de-seasoned data reported in Fig \ref{Selling_plots} and \ref{Selling_plots_II}.  The coefficients obtained are reported in Tab.  \ref{selling_DID_table_delta}.  As one can point out from the plots,  it is true that there is a pre-trend in the data before the event,  however as proved numerically in the Supporting Info,  the slope of this pre-trend is almost two order of magnitude lesser with respect to the slope in the lock-down T$_{1}$ and $T_{2}$.  As consequence this pre-trend,  compared to the lock-down effect,  can be considered negligible.   In order to make a further control,  a placebo test was performed by choosing timing before the pandemic of COVID19.  As illustrated in the Supporting Info,  this test was successful.  On the other side,  concerning $T_{3}$ only the effect for clothing and footwear can be considered significant,  also by considering the pre-trend slope as discussed in the Supporting Information.  Thus,  concerning $T_{3}$,  we considered not null,  in the scenario simulations, only the clothing and footwear sectors.  By inspecting the Tab.  \ref{selling_DID_table_delta} we see that the most severe damage of lock-down hit the clothing,  and the footwear.  On the other side the Household kids and the Household tools seems lesser affected by the lock-down.  The same DID regression was used  for the subsidies,  using as control group the extraordinary solidarity subsidies (cassa integrazione straordinaria solidarietà): the plot of the data and the DID coefficients are reported in Fig. \ref{CIG_sector_plots} and in Tab \ref{CIG_DID}.  Also in this case the placebo tests were successfully performed.  An interesting insight is also provided by the inspection of the extraordinary subsidies for renovation ( \textit{cassa integrazione ristrutturazione}): these subsidies,  as compared to the solidarity ones,  seems much more sensitive with respect to external shock in particular for trade sector

@@

>>>>

\title{\normalfont\spacedallcaps{The economic consequences of different attitudes of a policy maker: a combined epidemiological-econometric study}} % The article title

@@

{\centering
\includegraphics[width=0.5\textwidth]{Phase_transition.pdf}
\par
}

@@

The recent pandemic due to the spread of SARS-CoV-2 virus,  opened a highly debated issue about what it would be the best approach of the policy maker to face the epidemic.  Differently from the past pandemics and in particular of the XX century (e.g. Spanish Flu,  Asiatic Flu and Hong-Kong Flu),  for this pandemic a very large amount of data are easily accessible.  As a consequence not only the modelling of the virus diffusion but also the effect on socio-economic texture for different countries can be investigated with a finer resolution.  Among the different scientific challenges that can come up in this context,  an interesting one involves socio-economic effects of the attitude of the policy-maker (PM) to block the circulation of people (lock-down) in order to reduce the contagion rate (more formally the reproduction number, as described in Supporting Information).  Indeed the policy maker can adopt, at a first approximation,  a linear combination between these two extreme approaches: forcing all people to stay at home or to \textit{laissez-faire}.  In the first extreme,  the spread of virus is,  of course, stopped but,  on the other side,  the toll for such approach is that not only the economic activity (and so the income of people/firms) but also that the furniture of the primary goods are stopped.  On the other side,  if no lock-down measures are taken by the PM the toll to be paid will not only be the high number of deaths but also the economic damage produced by the very high number of deaths \cite{Correia2020,karlsson2014impact}.  In practice the PM can adopt intermediate approaches that shut down activities that contribute much more to diffusion with respect to other (for this purpose a very fine analysis was provided by Li et al.  in  \cite{li2021temporal} and by Brauner et al. in \cite{brauner2021inferring}).  As consequence the lock-down efficacy,  within certain limits,  can be tuned.  In the literature different scholars \cite{kabir2020evolutionary,rowthorn2020cost} challenged the issue of finding the optimal lock-down policy for minimizing the economic impact as well as the deaths.  In particular for the model in Ref. \cite{kabir2020evolutionary} it is assumed that the policy-maker knows perfectly the consequences of his/her choices and that he/she can act without delay to impose the optimal choice; finally it is assumed that the PM can impose a continuous factor for the lock-down while for different countries such factor seems to be much more discrete.  It can be argued that most of these drawbacks of this last formidable research, are entangled with the fact that a deterministic approach was considered for the activation of the lock-downs by the policy maker.  On this basis we would propose here an alternative way to model the decision of the policy maker that is based on a stochastic model instead of on a deterministic one.  Furthermore, differently with respect to the previous researches that focused basically on macroeconomic impact,  here we estimated and put in the model the impact of the different lock-down at microeconomic level: in particular,  by means of difference in difference,  we evaluated the effect of the different levels of lock-down on different sales sector as well as on subsidies (\textit{Cassa Integrazione Ordinaria}) in Italy.  Thus the final output of the model will be not only the cumulative deaths,  but also the economic damage for each selling sector.  Moreover here we also considered that there is not only an economic cost for each death as done by \cite{kabir2020evolutionary} but there is also average cost for each infected person (referring to Italian data),  because a consistent part of them may be recovered or even should take the intensive therapy.  As we will show by varying with continuity the probabilistic parameter by which the PM impose the lock-down,  a discontinuity in the SIRD model and in the days of the lock-down was obtained.  Such behaviour belongs to a class of phenomena that, in physical sciences, is called a phase transition.  For each phase we will discuss the result of the simulation and then we will compare them in order to get a general insight.

@@

The model of the present study is composed by an epidemiological part that shapes the diffusion of the virus and then its output is used by the economic model to quantify the damage. Thus we will discuss the epidemiological part and then the economic one

@@

Among the very large number of compartmental model available in the literature \cite{vynnycky2010introduction} we considered, as the simulator of the epidemic diffusion the simplest one: the Susceptible-Infectious-Recovered-Dead (SIRD).  Our choice is motivated by the fact that this relative simple model provides the gross features of an epidemic \cite{fernandez2020estimating,al2020forecasting} with a relative small number of parameters \footnote{One in principle can consider a SIRD model,in which the parameters that are time-dependent, as done by Ferrari et al. in Ref. \cite{ferrari2020modelling} for the description of Italian situation.  On the other side it is possible to increase the complexity of the model with other compartments as done in the following paper \cite{giordano2020modelling} by Giordano et al.  Note that in this last case the resolution of 9 differential equation is required (accompanied by the estimation of a large number of parameters)}). The SIRD model,first proposed by Kermack and McKendrick in 1927 \cite{kermack1927contribution},  is given by the following set of differential equations\cite{vynnycky2010introduction}:  

@@

known as the basic reproduction factor. This number is the average number of people that are infected by a single individual and describes if the epidemic is in a negative feedback (R$_{0}$<1),  stationary (R$_{0}$=1),  or in a positive feedback (R$_{0}$>1).  As consequence if the epidemic is in a negative feedback will be dissipate, while if it is in positive feedback will grow.  Note that in this simple model, since the parameters are not time dependent this factor is constant.  As performed by Ferrari et al. \cite{ferrari2020modelling} time dependent parameters are considered a time depended parameter called $R_{t}$.  For the present work,  we limited to constant parameters: in particular we considered the parameter estimation for Lombardy provided by Neves and Guerrero in Ref \cite{neves2020predicting}: $\beta$ was set equal to 0.55 while $\gamma$ equal to $\frac{1}{7}$. The overall population $N$ was set to 60M,  in order to simulate Italian population.  Within the daily temporal evolution of this model,  that was obtained by numerically solving the differential equation above via the \textit{DeSolve} package,  we considered a trigger activated by the PM every 7 days: if the number of infected people normalized by the overall population is more than $1\times 10^{-7}$,  there is a probability that the PM impose laws that multiply the $\beta$ factor by $0.7$,  if the normalized infected people are more than 10 the previous threshold he/she will impose with a certain probability restrictions that multiply $\beta$ starting $\beta$ factor by $0.25$;  finally,  if the  threshold is exceeded more than 50 times, the PM will impose with a certain probability restrictive measures that multiply $\beta$ by $0.025$.  These attenuation parameters were adjusted taking in consideration results of Marziano et al. in Ref. \cite{marziano2021retrospective}. As we said the PM acts with a certain probability, more formally stochastically: each week a random number (from zero to one) is extracted: if this is higher than a certain threshold, the relative restrictive decision is taken, otherwise not. The threshold value captures the PM attitude to impose the lock-down: lower values model a careful PM,  hight value a lazy one.  In this way the model is able to simulate different scenarios for the different PM attitude: as we will see this can produce two very different results.  At the end of the simulation,  beside the values given by the standard SIRD model (recovered and deaths),  the model also provide the number of days in which each restriction was active: these values are then used for the economic model in order to evaluate the economic effect due to the restrictions and PM strategy.  It is worth nothing that here as lock-down we considered only the national one that was applied in the first wave of the epidemics: from March 2020 to June 2020. This choice is motivated by the fact that modelling a unique system is more simpler with respect to modelling an ensemble of a communicating clusters that represent region or provinces: therefore,  if one is interested to model the second wave of pandemics such approach should be undertaken.  Moreover in this latter case,  as a further degree of complexity,  the economic data described in the next subsection must be at regional or province level,  and as far as the authors know, such data are not available.  For these reasons here the modelling will be always referred to national data and national lock-downs,  therefore the period after June when starting from October regional lock-down were imposed,  will be not considered.

@@

The economic impact for each epidemiological scenario is shaped as follow: a first set of parameters,  as the economic value of a death and of being infected by COVID-19, was taken directly from the reports/documentation of official sources; other parameters as the effect on selling for different areas and on the unemployment benefit (Cassa Integrazione Guadagni) were evaluated with empirical approaches from raw data.  Concerning the first set,  the number of deaths is multiplied by the maximum compensation value provided by the Court of Milan \cite{tribunaleMilano} for manslaughter (300k EUR). This choice is based on the idea that,  if the PM acts improperly,  can be incriminated for manslaughter (with the consent of parliament that has to validate the incrimination) and than,  if judged guilty,  charged by this amount for each death\footnote{In principle the judge also keep into account the age of the deaths: this in principle require an epidemiological model in which also the age of people is keep into account. In this case however a system of partial differential equation should be solved making the calculation and the computational cost incredibly high.}.  Beside this impact,  there is also the cost associated to the medical care of each ill people,  for this we considered the average value calculated by National Anti-Corruption Authority (ANAC) \cite{Anac}: 28.180 kEUR \footnote{It is worth noting that,  in principle,  there is also another import health-care impact due to the fact that the ill people for COVID-19 saturate the health system thus making it unreachable for other diseases.  This spillover translates in to more death and more ill people with respect to the baseline situation where there isn't an pandemic: however,  by now,  this effect is difficult to quantify and so we did not included in to the present model}.  Among the different sectors affected by the pandemic and the consequent lock-down,  we focused on the selling for the following ATECO-2007 \cite{Ateco2007} categories \footnote{In the rest part of the paper these categories will be referred as the part of the name labelled in blue. 
}: \textcolor{blue}{Food}, \textcolor{blue}{Clothing} and furs, \textcolor{blue}{Footwear}/leather and travel articles, \textcolor{blue}{Appliances}/radios/televisions and
tape recorders,\textcolor{blue}{Furniture}/textile articles/furnishings for the home, 
\textcolor{blue}{Photo-optics}/films/compact discs/ audio-video cassettes and musical instruments, Durable and non-durable \textcolor{blue}{Household kinds}, \textcolor{blue}{Household tools} and hardware,  \textcolor{blue}{Games}/toys/sports and camping articles. The choice to use sales as a parameter for the evaluation of the lock-down lies on the fact that with them  is possible to capture not only the contraction for each sector but also the loss for the public treasury due to to the reduced incomes from the VAT\footnote{For this purpose another sector that in principle can be also considered is the contraction of fuel selling, due to the reduced mobility, where in addition to the VAT there is also a fixed taxation (accisa).  Such calculation may be considered as a future outlook of this work}.  Beside the selling we also considered the unemployment benefit for the following ATECO sectors: \textcolor{blue}{Manufacturing} activities, \textcolor{blue}{Construction} , \textcolor{blue}{Wholesale} and retail trade/ repair of motor vehicles-motorcycles and personal and household goods.  In this case, the choice to use also this parameter is based on the fact that this is the first aid provided by the government for the firms that were damaged by the contraction in demand as well as by the fact that for different sector of them production was also banned by law during the lock-down \footnote{In principle one can be also interested to disentangle these two effects: in this case an interesting option is to study the second-wave of the epidemics in Italy that happened in the second half of the 2020. Contrary with respect to the first wave, the lock-down did not banned by law the production.  As told previously this option was not considered by the authors since the most of the economic data were available, with the temporal resolution considered in this work, only at national level.}.  For the empirical evaluation of the impact on selling and unemployment benefit we performed a multiple time Difference in Difference as used in Refs. \cite{draca2011panic} and \cite{wooldridge2012introductory}. The following regression was performed: 

@@

where $Y$ is the selcted outcome, $\alpha$ the intercept, $T_{i}$ a dummy variable for the lock-down timing i,  C a dummy for the treated group and $\epsilon$ a error term.  As control group for the sales,  we considered the food ones since,  in principle,  people can be considered to use almost the same amount of food regardless with respect to the lock-down \footnote{It is true that a slight increase of food consumption was done by people that were forced to stay at home as the plots in Fig.  \ref{Selling_plots} \ref{Selling_plots_II} captures: however as proven numerically in the Supporting Info this change does not significantly affect the DID estimations } (note that in this selling also the one for the restaurants are included: thus,  if it is true that the restaurants were closed,  it is also true that people will purchase basically the same amount of food from the supermarket that they will consume into restaurant if these were opened).  On the other side,  for subsidies \textit{Cassa integrazione},  we considered the Cassa Integrazione Straordinaria - Solidarità as the control group (note that this subsidy is different with respect Cassa Integrazione Solidarità that was dedicated to sectors not covered by Cassa Integrazione Ordinaria).  This subsidy can be used by firms,  damaged by the pandemics and by the lock-downs,  in order to reduce their labour cost but at same time guaranteeing to the workers part of their original salary.   During the first months of the pandemic firms,  following the rules stated by the Italian Governament (Decreto Cura Italia \cite{curaitalia}),  firms that would reduce their labour cost,  first forced the employees to use their holiday budget and than,  after it was run out,  they put the employees in to the Cassa Integrazione Ordinaria.  Thus the Cassa Integrazione Straordinaria-Solidarietà can be considered as not treated by the first wave of the lock-down while the ordinary one the treated.  The DID considered here has three different times for the national lock-down:  for the sales we considered,  as  done for the epidemiological model,  the months of March 2020 and May 2020 as medium lock-down, the month of April 2020 as high lock-down and June 2020 as weak lock-down.  On the other side concerning the subsidies,  we considered as medium lock-down only May 2020: such choice was motivated by the fact that the effect of firms to use the subsidies for workers was slightly delayed basically due to the fact that firms forced their employees first to use their holiday and then the \textit{Cassa Integrazione Ordinaria}: as consequence,  because of this buffer effect, in the month of March there is no significant effect of this subsidies although there is a significant reduction of hours worked (see e.g.  \cite{istat_lavoro}).  Performing the DID we estimated the coefficients for the effect of each intensity of lock-down on sales and subsidies,  then we we rescaled them for a week.  Finally we multiplied the number obtained from the previous calculation by the number of lock-down weeks,  with the respective intensity,  in each scenario simulated via the epidemiological model described before.  As consequence we obtained,  for each scenario,  an economics simulation based on parameters obtained from an empirical evaluation.

@@

The monthly sales data were take retrieved from the website of National Institute of Statistics \cite{istat}.  In particular we considered the period starting from 2018 up to June 2020. The choice to not consider the months after June and in particular the last part of the year lies on the fact that in the latter the lock-down were imposed at regional level and not to national level \footnote{With the only exception of Christmas holiday}  \footnote{In principle one may ask why the present analysis was not performed on regional cluster making it more flexible: unfortunately the economic data used here, as far as the author knows,  were not available,  at all for regional cluster.  Moreover if regional cluster were considered it was necessary to model an ensemble of SIRD model that communicate each other with a defined rate (that change also with respect to the lock-down restriction).  This make the model much more complicated.  However if all the data that are necessary for performing the analysis will become available the author may consider, as an outlook, to extend the present analysis to regional clusters}.  As discussed in Supporting Information,  these data were not de-seasoned: thus we performed a de-seasoning via the \textit{Forecast} R package \cite{forecast} that uses a Hilbert-Huang transform  \cite{huang1998empirical} for the decomposition of a time series data.  Concerning the subsidies we retried the data from the \textit{Osservatorio Cassa integrazione guadagni e fondi di solidarietà} on the \textit{Istituto nazionale della previdenza sociale} webpage \cite{inps}.  It is worth nothing that here we considered only the authorized subsidy and not the asked one.  Furthermore these data,  differently with respect to the sales one were not affected by seasonality noise,  and thus no de-seasoning was necessary.  We checked also if any significant effect on DID coefficient was obtained for sales and for subsidies by adding one and two year before 2018 (e.g. 2016 and 2017): no significant effect was found.

@@

As done for the section Model and methods we will divide the discussion of the results in the following way: first the outcomes of the economic model will be presented,  then on the basis of them,  we will discuss the  scenarios obtained with them via the epidemiological model.  Finally we will discuss the overall results.

@@

For each sales category we run the regression reported in the Eq.  \ref{DID-eq} on de-seasoned data reported in Fig \ref{Selling_plots} and \ref{Selling_plots_II}.  The coefficients obtained are reported in Tab.  \ref{selling_DID_table_delta}.  As one can point out from the plots,  it is true that there is a pre-trend in the data before the event,  however as proved numerically in the Supporting Info,  the slope of this pre-trend is almost two order of magnitude lesser with respect to the slope in the lock-down T$_{1}$ and $T_{2}$.  As consequence this pre-trend,  compared to the lock-down effect,  can be considered negligible.   In order to make a further control,  a placebo test was performed by choosing timing before the pandemic of COVID19.  As illustrated in the Supporting Info,  this test was successful.  On the other side,  concerning $T_{3}$  (low lock-down), only the effect for clothing and footwear can be considered significant,  also by considering the pre-trend slope as discussed in the Supporting Information.  Thus,  concerning $T_{3}$,  we considered not null,  in the scenario simulations,  only the clothing and footwear sectors.  By inspecting the Tab.  \ref{selling_DID_table_delta} we see that the most severe damage of lock-down hit the clothing,  and the footwear.  On the other side the Household kids and the Household tools seems lesser affected by the lock-down.  The same DID regression was used  for the subsidies,  using as control group the extraordinary solidarity subsidies (cassa integrazione straordinaria solidarietà): the plot of the data and the DID coefficients are reported in Fig. \ref{CIG_sector_plots} and in Tab \ref{CIG_DID}.  Also in this case the placebo tests were successfully performed.  An interesting insight is also provided by the inspection of the extraordinary subsidies for renovation ( \textit{cassa integrazione ristrutturazione}): these subsidies,  as compared to the solidarity ones,  seems much more sensitive with respect to external shock in particular for trade sector.

@@

\begin{figure}
 \centering
 \includegraphics[width=\textwidth]{Figures/Phase_transition.pdf}
\caption{The number of deaths of SIRD scenario with stochasic lockdown as obtained by chaing the PM attitute to active the lockdown (e.g.  by modifying the probability parameter by which the lock-down is imposed).  As the PM attitude is near 1.3 a sharp discontinuity is present in the overall death: thus a different phase is obtained.  It is worth nothing the similarity of this plot with the gold-standard one of phase transition: the Ising model (see \cite{Ising} or \cite{Ising_2} for a more deep analysis)} 
\label{Phase_transition}
\end{figure}
